Reactive SINDy: Discovering governing reactions from concentration data

### 2018-12-11 docs/resubmission_jcp/cover_letter/resubmission_cover_letter.tex
resubmission preparations

\documentclass[11pt,english]{article}
\usepackage[T1]{fontenc}
\usepackage{microtype}
\usepackage[utf8]{inputenc}
\usepackage{babel}
\usepackage{hyperref}
\setlength\parindent{0pt}

@@

\title{}

@@

\textbf{{\Large Response to the reviewers}}\newline

\textbf{{\large Reviewer \# 2}}\newline

@@

\textbf{Answer:}
We now applied the method to a MAPK cascade as well as a predator-prey model that contains
temporal bimodality and show that we can recover the correct reaction
network with the correct regularization hyperparameters.

>>>>

\documentclass[11pt,english]{scrartcl}
\usepackage[T1]{fontenc}
\usepackage{microtype}
\usepackage[utf8]{inputenc}
\usepackage{babel}
\usepackage{hyperref}
\setlength\parindent{0pt}

@@

@@

\begin{flushleft}
    \large
    \textbf{Re-submission of manuscript:} Reactive SINDy: Discovering governing reactions from concentration data
\end{flushleft}
\begin{flushright}
    \today
\end{flushright}
~\\

Dear editorial board members,\\

please find attached our revised manuscript in tex format and another compiled pdf document highlighting the changes. Further we give the response to the reviewers below.\\

\textbf{{\large Response to Reviewer \#1}}\newline

Thank you for the review.\newline

\textbf{{\large Response to Reviewer \#2}}\newline

@@

\textbf{Answer:}
We now applied the method to a MAPK cascade as well as a predator-prey model that contains
temporal bimodality and show that we can recover the correct reaction
network with the correct regularization hyperparameters.\newline

Best regards,\\

Moritz Hoffmann, Christoph Fröhner, and Frank Noé

### 2018-12-11 docs/resubmission_jcp/manuscript.tex
resubmission preparations

We additionally demonstrate the algorithm
on time series data of the mitogen activated protein kinases (MAPK) pathway as 
an example for a bimodal system and on time series data of the
Lotka--Volterra system which describes oscillatory predator-prey dynamics subject to
social friction. In both systems reactive SINDy recovers the generating 
reaction network whereas non-sparse estimation detects many false processes.

@@

\begin{figure}[h!]
\centering \includegraphics[width=\singlecolumnwidth]{figures_jcp/1.pdf}
\caption{\label{fig:network}The regulation network example described in Sec.~\ref{sec:results}.
Each circle depicts a species, each arrow corresponds to one reaction.
Blue arrows denote transcription from $\mathrm{DNA}$ to $\mathrm{mRNA}$,
green arrows denote translation from $\mathrm{mRNA}$ to protein,
and red arrows denote the regulatory network.}
\end{figure}
\begin{table}
\centering%
\ifthisispreprint
% ignore
\else
\resizebox{.95\columnwidth}{!}{%
\fi
\begin{tabular}{cccccccccc}
 & $\mathrm{DNA}_{\mathrm{A}}$ & $\mathrm{mRNA}_{\mathrm{A}}$ & $\mathrm{A}$ & $\mathrm{DNA}_{\mathrm{B}}$ & $\mathrm{mRNA}_{\mathrm{B}}$ & $\mathrm{B}$ & $\mathrm{DNA}_{\mathrm{C}}$ & $\mathrm{mRNA}_{\mathrm{C}}$ & $\mathrm{C}$\tabularnewline
\midrule
\textbf{(a)} & $1$ & $2$ & $0$ & $1$ & $0$ & $3$ & $1$ & $0$ & $0$\tabularnewline
\textbf{(b)} & $1$ & $1.5$ & $0$ & $1$ & $0$ & $2$ & $1$ & $0$ & $1$\tabularnewline
\end{tabular}
\ifthisispreprint
% ignore
\else
}
\fi
\caption{\label{tab:Initial-conditions}Initial conditions \textbf{(a)} and \textbf{(b)} used to generate
concentration time series. Reaction rates can be found in Tab.~\ref{tab:reaction-library}.}
\end{table}
The proposed estimation method is applied to analyze these time series
of concentrations in order to recover the underlying reaction network
from data. To this end we use the library of ansatz functions given
in Tab.~\ref{tab:reaction-library}, which contains a large number
of possible reactions, only few of which are actually part of the
model.

@@

We first set the hyperparameter $\alpha=0$ in the minimization problem
(\ref{method:minimizationproblem}), which results in constrained
least-squares regression without any of the regularization terms.
In this case we estimate a reaction network that can reproduce the
observations almost exactly (Fig.~\ref{fig:case-1-lsq}). However,
the result is mechanistically wrong as the sparsity pattern does not
match the reaction network used to generate the data. On the one hand
many spurious reactions are estimated that were not in the true reaction
scheme and would lead to wrong conclusions about the mechanism, such
as $\mathrm{A}+\mathrm{A}\rightharpoonup\mathrm{A}$ and $\mathrm{A}+\mathrm{C}\rightharpoonup\mathrm{C}$.
More dramatically, the reaction responsible for the decay of $\mathrm{A}$
particles is completely ignored (Fig.~\ref{fig:case-1-sparsity-pattern}).
\begin{figure}
\centering\includegraphics[width=\singlecolumnwidth]{figures_jcp/2.pdf}\caption{\label{fig:case-1-lsq}Concentration time series generated from integrating
the reaction network shown in Fig.~\ref{fig:network}a. The initial
condition prescribes positive concentration values only for $\mathrm{B}$
protein and $\mathrm{mRNA}_{\mathrm{A}}$ species (Tab.~\ref{tab:Initial-conditions}a).
This initial condition is used in the subsequent sections for further
analysis. Gray dots depict concentration time series yielded from
the LSQ rates estimated in Sec~\ref{sec:case-1}.}
\end{figure}

@@

\begin{figure*}
\centering \includegraphics[width=0.82\doublecolumnwidth]{figures_jcp/3.pdf}
\caption{\label{fig:case-1-sparsity-pattern}Estimated reaction rates in the
system described in Sec.~\ref{sec:case-1}. The y and x axes contain
reaction educts and products, respectively. A circle at position $(i,j)$
represents a reaction $i\rightharpoonup j$ whose rate has a linear
relation with the area of the circle. The black outlines denote the
reactions with which the system was generated and contain the respective
rate value. Red crosses denote reactions that were used as additional
ansatz reactions. Blue circles are estimated by LSQ and orange circles
depict rates which were obtained by solving the minimization problem
(\ref{method:minimizationproblem}). The latter rates are subject
to a cutoff $\kappa=0.22$ corresponding to the green circle's area
under which a sparse solution with the correct processes can be recovered.
If a certain rate was estimated in both cases, two wedges instead
of one circle are displayed.}
\end{figure*}
%\ifthisispreprint
% ignore
%\else
%\twocolumngrid
%\fi

@@

\begin{figure}
\centering \includegraphics[width=\singlecolumnwidth]{figures_jcp/4.pdf}
\caption{\label{fig:case2-convergence}Convergence of the estimation error
when estimating the system described in Sec.~\ref{sec:case-1} with
varying levels of noise by application of reactive SINDy (\ref{method:minimizationproblem})
with and without regularization in blue and orange, respectively.
The procedure was independently repeated $10$ times with different
realizations giving rise to the mean and standard deviation depicted
by solid lines and shaded areas, respectively. \textbf{(a)}: The number
of detected spurious reactions up to the cutoff value introduced in
Sec.~\ref{sec:case-1} over different levels of noise. \textbf{(b)}:
The estimation error given by the mean absolute error between the
generating reaction rates $\xi$ and the estimated reaction rates
$\hat{\xi}$ over different levels of noise.}
\end{figure}
The hyperparameters $(\alpha,\lambda)$ are obtained by shuffling
the data and performing a $10$-fold cross validation.

@@

\begin{figure}
\centering \includegraphics[width=\singlecolumnwidth]{figures_jcp/5.pdf}
\caption{\label{fig:case3-convergence}Convergence of estimation error of reaction
schemes from noisy gene-regulation data starting from two different
initial conditions under decreasing levels of noise. The minimization
problem (\ref{method:minimizationproblem}) was solved for $\alpha=0$
(LSQ) and with regularization. This was repeated $10$ times on different
sets of observation data generated by Gillespie SSA, giving rise to
mean and standard deviation (solid lines and shaded areas, respectively).
\textbf{(a)}: Concentration time series corresponding to the initial
conditions, generated by integrating the reaction-rate equations.
The first initial condition is identical to the one used in Sec.~\ref{sec:case-1}
and Sec.~\ref{sec:case-2}. The second initial condition (Tab.~\ref{tab:Initial-conditions}b)
prescribes positive initial concentrations for $\mathrm{mRNA}_{\mathrm{A}}$,
$\mathrm{B}$, and $\mathrm{C}$ species. The gray graphs are sample
realizations of integration using the Gillespie SSA. \textbf{(b)},\textbf{(c)}:
Analogously to Fig.~\ref{fig:case2-convergence} with the difference
that 20-fold cross validation was used for hyperparameter estimation.}
\end{figure}
The hyperparameters $(\alpha,\lambda)$ are obtained by shuffling
the data and performing a $20$-fold cross validation.

@@

\begin{figure}
    \centering
    \includegraphics[width=\singlecolumnwidth]{figures_jcp/6.pdf}
    \caption{\label{fig:mapk}Application of reactive SINDy to the MAPK pathway system. \textbf{(a)}~The response curve of the MAPK cascade as a function of external stimulus given as a constant concentration $[\mathrm{S}]$. The activity is the steady state concentration of activated transcription factors $[\mathrm{TF*}]$. Dashed lines show the values of $[\mathrm{S}]$ at which concentration time series data was generated. \textbf{(b)}~Estimated rate coefficients of candidate reactions (see Tab.~\ref{tab:reaction-library-mapk}) after application of reactive SINDy (regularized) to the time series data. Least-squares estimation (LSQ) and the ground truth model for comparison.}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=\singlecolumnwidth]{figures_jcp/7.pdf}
    \caption{\label{fig:lv}Application of reactive SINDy to the Lotka--Volterra system with social friction. \textbf{(a)}~Concentration data as a function of time for predator and prey species. \textbf{(b)}~Estimated rate coefficients of candidate reactions (see Tab.~\ref{tab:reaction-library-lv}) after application of reactive SINDy (regularized) to the time series data. Least-squares estimation (LSQ) estimation and the ground truth model for comparison.}
\end{figure}

@@

\centering\includegraphics[width=\singlecolumnwidth]{figures_jcp/8.pdf}\caption{\label{fig:reaction-rates-sorted}Reaction rates sorted by their magnitude
to determine the cutoff $\kappa=0.22$ of Sec.~\ref{sec:case-1}.
The rates were estimated using the regularized minimization problem.}

>>>>

We additionally demonstrate the algorithm
on time series data of the mitogen activated protein kinases (MAPK) pathway as 
an example for a bimodal system and on time series data of the
Lotka--Volterra system which describes oscillatory predator-prey dynamics subject to
social friction. In both systems reactive SINDy recovers the generating 
reaction network whereas non-sparse estimation detects many spurious processes.

@@

\begin{figure}[h!]
\centering \includegraphics[width=\singlecolumnwidth]{1.pdf}
\caption{\label{fig:network}The regulation network example described in Sec.~\ref{sec:results}.
Each circle depicts a species, each arrow corresponds to one reaction.
Blue arrows denote transcription from $\mathrm{DNA}$ to $\mathrm{mRNA}$,
green arrows denote translation from $\mathrm{mRNA}$ to protein,
and red arrows denote the regulatory network.}
\end{figure}
\begin{table}
\centering%
\ifthisispreprint
% ignore
\else
\resizebox{.95\columnwidth}{!}{%
\fi
\begin{tabular}{cccccccccc}
 & $\mathrm{DNA}_{\mathrm{A}}$ & $\mathrm{mRNA}_{\mathrm{A}}$ & $\mathrm{A}$ & $\mathrm{DNA}_{\mathrm{B}}$ & $\mathrm{mRNA}_{\mathrm{B}}$ & $\mathrm{B}$ & $\mathrm{DNA}_{\mathrm{C}}$ & $\mathrm{mRNA}_{\mathrm{C}}$ & $\mathrm{C}$\tabularnewline
\midrule
\textbf{(a)} & $1$ & $2$ & $0$ & $1$ & $0$ & $3$ & $1$ & $0$ & $0$\tabularnewline
\textbf{(b)} & $1$ & $1.5$ & $0$ & $1$ & $0$ & $2$ & $1$ & $0$ & $1$\tabularnewline
\end{tabular}
\ifthisispreprint
% ignore
\else
}
\fi
\caption{\label{tab:Initial-conditions}Initial conditions \textbf{(a)} and \textbf{(b)} used to generate
concentration time series. Reaction rates can be found in Tab.~\ref{tab:reaction-library}.}
\end{table}
The proposed estimation method is applied to analyze these time series
of concentrations in order to recover the underlying reaction network
from data. To this end we use the library of ansatz functions given
in Tab.~\ref{tab:reaction-library}, which contains a large number
of possible reactions, only few of which are actually part of the
model.

@@

We first set the hyperparameter $\alpha=0$ in the minimization problem
(\ref{method:minimizationproblem}), which results in constrained
least-squares regression without any of the regularization terms.
In this case we estimate a reaction network that can reproduce the
observations almost exactly (Fig.~\ref{fig:case-1-lsq}). However,
the result is mechanistically wrong as the sparsity pattern does not
match the reaction network used to generate the data. On the one hand
many spurious reactions are estimated that were not in the true reaction
scheme and would lead to wrong conclusions about the mechanism, such
as $\mathrm{A}+\mathrm{A}\rightharpoonup\mathrm{A}$ and $\mathrm{A}+\mathrm{C}\rightharpoonup\mathrm{C}$.
More dramatically, the reaction responsible for the decay of $\mathrm{A}$
particles is completely ignored (Fig.~\ref{fig:case-1-sparsity-pattern}).
\begin{figure}
\centering\includegraphics[width=\singlecolumnwidth]{2.pdf}\caption{\label{fig:case-1-lsq}Concentration time series generated from integrating
the reaction network shown in Fig.~\ref{fig:network}a. The initial
condition prescribes positive concentration values only for $\mathrm{B}$
protein and $\mathrm{mRNA}_{\mathrm{A}}$ species (Tab.~\ref{tab:Initial-conditions}a).
This initial condition is used in the subsequent sections for further
analysis. Gray dots depict concentration time series yielded from
the LSQ rates estimated in Sec~\ref{sec:case-1}.}
\end{figure}

@@

\begin{figure*}
\centering \includegraphics[width=0.82\doublecolumnwidth]{3.pdf}
\caption{\label{fig:case-1-sparsity-pattern}Estimated reaction rates in the
system described in Sec.~\ref{sec:case-1}. The y and x axes contain
reaction educts and products, respectively. A circle at position $(i,j)$
represents a reaction $i\rightharpoonup j$ whose rate has a linear
relation with the area of the circle. The black outlines denote the
reactions with which the system was generated and contain the respective
rate value. Red crosses denote reactions that were used as additional
ansatz reactions. Blue circles are estimated by LSQ and orange circles
depict rates which were obtained by solving the minimization problem
(\ref{method:minimizationproblem}). The latter rates are subject
to a cutoff $\kappa=0.22$ corresponding to the green circle's area
under which a sparse solution with the correct processes can be recovered.
If a certain rate was estimated in both cases, two wedges instead
of one circle are displayed.}
\end{figure*}
%\ifthisispreprint
% ignore
%\else
%\twocolumngrid
%\fi

@@

\begin{figure}
\centering \includegraphics[width=\singlecolumnwidth]{4.pdf}
\caption{\label{fig:case2-convergence}Convergence of the estimation error
when estimating the system described in Sec.~\ref{sec:case-1} with
varying levels of noise by application of reactive SINDy (\ref{method:minimizationproblem})
with and without regularization in blue and orange, respectively.
The procedure was independently repeated $10$ times with different
realizations giving rise to the mean and standard deviation depicted
by solid lines and shaded areas, respectively. \textbf{(a)}: The number
of detected spurious reactions up to the cutoff value introduced in
Sec.~\ref{sec:case-1} over different levels of noise. \textbf{(b)}:
The estimation error given by the mean absolute error between the
generating reaction rates $\xi$ and the estimated reaction rates
$\hat{\xi}$ over different levels of noise.}
\end{figure}
The hyperparameters $(\alpha,\lambda)$ are obtained by shuffling
the data and performing a $10$-fold cross validation.

@@

\begin{figure}
\centering \includegraphics[width=\singlecolumnwidth]{5.pdf}
\caption{\label{fig:case3-convergence}Convergence of estimation error of reaction
schemes from noisy gene-regulation data starting from two different
initial conditions under decreasing levels of noise. The minimization
problem (\ref{method:minimizationproblem}) was solved for $\alpha=0$
(LSQ) and with regularization. This was repeated $10$ times on different
sets of observation data generated by Gillespie SSA, giving rise to
mean and standard deviation (solid lines and shaded areas, respectively).
\textbf{(a)}: Concentration time series corresponding to the initial
conditions, generated by integrating the reaction-rate equations.
The first initial condition is identical to the one used in Sec.~\ref{sec:case-1}
and Sec.~\ref{sec:case-2}. The second initial condition (Tab.~\ref{tab:Initial-conditions}b)
prescribes positive initial concentrations for $\mathrm{mRNA}_{\mathrm{A}}$,
$\mathrm{B}$, and $\mathrm{C}$ species. The gray graphs are sample
realizations of integration using the Gillespie SSA. \textbf{(b)},\textbf{(c)}:
Analogously to Fig.~\ref{fig:case2-convergence} with the difference
that 20-fold cross validation was used for hyperparameter estimation.}
\end{figure}
The hyperparameters $(\alpha,\lambda)$ are obtained by shuffling
the data and performing a $20$-fold cross validation.

@@

\begin{figure}
    \centering
    \includegraphics[width=\singlecolumnwidth]{6.pdf}
    \caption{\label{fig:mapk}Application of reactive SINDy to the MAPK pathway system. \textbf{(a)}~The response curve of the MAPK cascade as a function of external stimulus given as a constant concentration $[\mathrm{S}]$. The activity is the steady state concentration of activated transcription factors $[\mathrm{TF*}]$. Dashed lines show the values of $[\mathrm{S}]$ at which concentration time series data was generated. \textbf{(b)}~Estimated rate coefficients of candidate reactions (see Tab.~\ref{tab:reaction-library-mapk}) after application of reactive SINDy (regularized) to the time series data. Least-squares estimation (LSQ) and the ground truth model for comparison.}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=\singlecolumnwidth]{7.pdf}
    \caption{\label{fig:lv}Application of reactive SINDy to the Lotka--Volterra system with social friction. \textbf{(a)}~Concentration data as a function of time for predator and prey species. \textbf{(b)}~Estimated rate coefficients of candidate reactions (see Tab.~\ref{tab:reaction-library-lv}) after application of reactive SINDy (regularized) to the time series data. Least-squares estimation (LSQ) estimation and the ground truth model for comparison.}
\end{figure}

@@

\centering\includegraphics[width=\singlecolumnwidth]{8.pdf}\caption{\label{fig:reaction-rates-sorted}Reaction rates sorted by their magnitude
to determine the cutoff $\kappa=0.22$ of Sec.~\ref{sec:case-1}.
The rates were estimated using the regularized minimization problem.}
